Dark Control:\\ Towards a Unified Account of Default Mode Function\\ by Markov Decision Processes

### 2017-12-06 paper/paper.tex
Corrections of the memory/levels edits for reviewer 2

We argue that DMN implication in many
advanced cognitive processes in humans
can be recast as prediction error minimization
based on probabilistic mental simulations
- ``covert forms of action and perception'' \citep{pezzulo2011grounding} -,
thus maximizing action outcome across multiple time scales.
Such a purposeful optimization objective
may be solved by a stochastic approximation
based on a brain implementation of Markov Chain Monte Carlo (MCMC) sampling
\citep{tenenbaum2011grow}.
Even necessarily imperfect memory recall,
random day-time mind-wandering, and
seemingly arbitrary dreams during sleep
may provide blocks of pseudo-experience to iteratively
optimize the behavior of the organism.

Evidence from computational modeling of human behavior \citep{kording2004bayesian} and cell recording experiments in ferrets \citep{fiser2004small} suggest that
the human brain is largely dedicated to
``the development and maintenance of [a]
probabilistic model of anticipated events''
\citep{raichle2005intrinsic}.
The present paper proposes a
process model that satisfies this contention.
We also contribute to the discussion of DMN function
by providing some of the first empirical evidence that
morphological variability in DMN regions
is linked to the reward circuitry (Fig. 2).
Finally, we detail how our process model relates
to previous accounts of DMN function and
we derive explicit hypotheses to be tested in future neuroscience
experiments (Box 2).
At this stage, it is important to clarify how different levels of observation
are at play in the current account. A process model is not solely intended to 
capture behavior of the agent but also the neurocomputational specifics of the agent.
In the following, we will use the word ``inference`` when describing the formal aspects
of the statistical model, the word ``prediction`` when discussing the neural implementation,
and words like ``forecast`` or ``forsee`` when refering to the behavior of the agent.

@@

The environment is generally taken as \textit{stochastic},
that is, changing in random ways. In addition, the environment is only
\textit{partially observable} in the sense that only limited aspects of the environment's
state are accessible to the agent's sensory perception.
\citep{starkweather2017dopamine}.
We assume that volatility of the environment
is realistic in a computational model which sets out
to explain DMN functions of the human brain.
% XXX: the following passage is already in the introduction
% Indeed, it is known that the more the external world is predictable,
% the more mental activity becomes detached from the actual sensory inputs
% \citep{antrobus1966studies, pope1978regulation},
% the more stimulus-independent thoughts occur, and
% the higher the neural activity in the DMN.
% Conversely, the more the currently executed task is unknown and unpracticed,
% the lower the DMN activity
% \citep{filler1973daydreaming, teasdale1995stimulus}.
% %
% These ``offline'' processes may however contribute to
% optimizing control of the organism.
% Such a framework of DMN funtion based on reinforcement learning
% can naturally embed human behavior
% in the tension between exploitative action with immediate gains and
% explorative action with longer-term reward schedules
% \citep{dayan2008decision}.
We argue that a functional account of the DMN based on RL
can naturally embed human behavior
in the tension between exploitative action with immediate gains and
explorative action with longer-term reward outcomes
\citep{dayan2008decision}.
In short, DMN implication in a diversity of
particularly advanced cognitive processes
can be parsimoniously explained as probabilistic mental
scene of experience simulations coupled with prediction error minimization
to calibrate action trajectories for
reward outcome maximization at different time scales.
Such a purposeful optimization objective
may be solved by a stochastic approximation
based on a brain implementation of MCMC sampling
\citep{tenenbaum2011grow}.

@@

% , defining $s(t)$, the state at time $t$, as the position in state space.
The process has $memory$ if the next state depends not only on the current state
but also on a finite number of past states.
Rational probabilistic planning can thus be reformulated
as a standard memoryless Markov process by simply expanding the
definition of the state $s$ to include experience episodes of the past.
This adds the capacity for memory, since then, the next state depends not only on the current state but also on a finite number of past states. This is precisely what Partially Observable MDPs (POMDPs) do ~\citep{starkweather2017dopamine,o2006making}.
Nevertheless, this mathematical procedure in POMDPs only accounts for implicit $memory$. Since the current paper is aiming at both plausibility at the behavioral and neural level, we will address later in the paper how our model captures the neurophysiological constraints of the DMN and can also account for the explicit part of $memory$ from a human agent perspective.

@@

\paragraph*{On implicit and explicit memory in the current model.}
Markov Processes are memoryless but this is possible mathematically to artificially insert the previous states of such model into the current state. While this may partially account for implicit memory at the behavioral level, it does not explain the underlying implementation at the neurophysiological level nor accounts for the explicit part of memory.
The implicit memory-based processing is already present in our MDP account of DMN function; it occurs in 3 three different forms: a) the action policy, b) the value function $\textemdash$as both of these are a result/product of the past$\textemdash$ and c) the deep architecture with many non-linear links in the hierarchical connections as realized by biological neural networks. Indeed, we reason about the brain has deep neural networks that present an implicit form of information compression obtained from life experience. This memory traces are implicitly stored in the neural machinery and can be retrieve as a form of Knowledge by simulation of action rather than accessed as a stored explicit representation \citep{pezzulo2011grounding}. The neural processes in the hippocampus can be seen as MCMC sampling from a formal statistical level but those memory recall are also the basis for scene construction processes across time-scales such as in the semantic hypothesis \citep{schacter2007remembering,Axelrod2017}. In a complementary way, neural encoding of reward information is not limited to value but incorporates specific information about the identity of the reward outcome, the predictive stimulus, or the action necessary to obtain the reward \citep{kahnt2017decade}.

>>>>

We argue that DMN implication in many
advanced cognitive processes in humans
can be recast as prediction error minimization
based on probabilistic mental simulations
- ``covert forms of action and perception'' \citep{pezzulo2011grounding} -,
thus maximizing action outcome across multiple time scales.
Such a purposeful optimization objective
may be solved by a stochastic approximation
based on a brain implementation of Markov Chain Monte Carlo (MCMC) sampling.
Even necessarily imperfect memory recall,
random day-time mind-wandering, and
seemingly arbitrary dreams during sleep
may provide blocks of pseudo-experience to iteratively
optimize the behavior of the organism.

Evidence from computational modeling of human behavior \citep{kording2004bayesian} and cell recording experiments in ferrets \citep{fiser2004small} suggest that
the human brain is largely dedicated to
``the development and maintenance of [a]
probabilistic model of anticipated events''
\citep{raichle2005intrinsic}.
The present paper proposes a
process model that satisfies this contention.
We also contribute to the discussion of DMN function
by providing some of the first empirical evidence that
morphological variability in DMN regions
is linked to the reward circuitry (Fig. 2).
Finally, we detail how our process model relates
to previous accounts of DMN function and
we derive explicit hypotheses to be tested in future neuroscience
experiments (Box 2).
At this stage, we emphasize the importance of differentiating which levels of observation are at play in the present account. A process model is not solely intended to capture behavior of the agent, such as cognitive accounts of DMN function, but also the neurocomputational specifics of the agent. Henceforth, we will use ``inference`` when describing aspects of the statistical model, ``prediction`` when referring to the neurobiological implementation, and words like ``forecast`` or ``forsee`` when referring to the behavior of the agent.

@@

The environment is generally taken as \textit{stochastic},
that is, changing in random ways. In addition, the environment is only
\textit{partially observable} in the sense that only limited aspects of the environment's
state are accessible to the agent's sensory perception.
\citep{starkweather2017dopamine}.
We assume that volatility of the environment
is realistic in a computational model which sets out
to explain DMN functions of the human brain.
% XXX: the following passage is already in the introduction
% Indeed, it is known that the more the external world is predictable,
% the more mental activity becomes detached from the actual sensory inputs
% \citep{antrobus1966studies, pope1978regulation},
% the more stimulus-independent thoughts occur, and
% the higher the neural activity in the DMN.
% Conversely, the more the currently executed task is unknown and unpracticed,
% the lower the DMN activity
% \citep{filler1973daydreaming, teasdale1995stimulus}.
% %
% These ``offline'' processes may however contribute to
% optimizing control of the organism.
% Such a framework of DMN funtion based on reinforcement learning
% can naturally embed human behavior
% in the tension between exploitative action with immediate gains and
% explorative action with longer-term reward schedules
% \citep{dayan2008decision}.
We argue that a functional account of the DMN based on RL
can naturally embed human behavior
in the tension between exploitative action with immediate gains and
explorative action with longer-term reward outcomes
\citep{dayan2008decision}.
In short, DMN implication in a diversity of
particularly advanced cognitive processes
can be parsimoniously explained as probabilistic mental
scene of experience simulations coupled with prediction error minimization
to calibrate action trajectories for
reward outcome maximization at different time scales.
Such a purposeful optimization objective
may be solved by a stochastic approximation
based on a brain implementation of MCMC sampling.

@@

% , defining $s(t)$, the state at time $t$, as the position in state space.
The process has $memory$ if the next state depends not only on the current state but also on a finite number of past states. Rational probabilistic planning can thus be reformulated as a standard memoryless Markov process by simply expanding the definition of the state $s$ to include experience episodes of the past. This adds the capacity for memory, since then, the next state depends not only on the current state but also on a finite number of past states. This is precisely what Partially Observable MDPs (POMDPs) aim to realize ~\citep{starkweather2017dopamine,o2006making}.
Nevertheless, this mathematical property in POMDPs mostly accounts for implicit $memory$. Since the current paper is focused on both plausibility at the behavioral and neurobiological level, we will address below how our model captures the neurophysiological constraints of the DMN and can also account for the explicit part of $memory$ characteristic for human agents.

@@

\paragraph*{On implicit and explicit memory in the current model.}
Markov Processes are memoryless but this it is mathematically feasible to insert the previous states of such model into the current state. While this may partially account for implicit memory at the behavioral level, it may not explain the underlying implementation at the neurophysiological level nor accounts for the explicit part of memory. Implicit memory-based processing arises in our MDP account of DMN function in three different forms: successive updates of a) the action policy and b) the value function, both being products of the past and c) the deep non-linear links in the hierarchical connections realized by biological neural networks. Indeed, we reason about the brain has deep neural networks that present an implicit form of information compression obtained from life experience. Such memory traces are implicitly stored in the neural machinery and can be retrieve as a form of knowledge during simulation of action rather than accessed as a stored explicit representation \citep{pezzulo2011grounding}. The neural processes in the hippocampus can be seen as MCMC sampling from a formal statistical level but those memory recall are also the basis for scene construction processes across time-scales such as in the semantic hypothesis \citep{schacter2007remembering,Axelrod2017}.
